A fast algorithm for global code motion of congruent instructions

### 2017-11-11 cgo/paper.tex
sinking live range

%% Title information
\title[SSA GCM]{Global code motion of congruent instructions}         %% [Short Title] is optional;

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a \gcm{} (\GCM{}) compiler optimization which schedules congruent
instructions across the program.  Not only \GCM{} saves code size, it exposes
redundancies as well, it exposes more instruction level parallelism in the
basic-block to which instructions are moved, and it enables other passes like
loop invariant motion to remove redundancies. The cost model to drive the code
motion is based on liveness analysis on \SSA{} representation such that the
(virtual) register pressure does not increase resulting in 2\% fewer spills on
the \SPEC{} benchmark suite.

@@

\section{A fast algorithm for code motion congruent computations}
The algorithm factors out the reachability of values such that multiple queries
to find reachability of values are fast. This is based on finding the dominance
frontiers and the post-dominance frontiers (PDF) of the CFG. These points, being
structural properties of a graph, do not change during code-motion. A
post-dominance frontier is the basic block where anticipability (ANTIC) of an
instruction (I) is checked because it is the basic block on which the basic
block containing I is control dependent.

So we introduce a data structure ($\chi$ nodes) to keep track of values flowing out
of a basic block with more than one successor. We only do this for values with
multiple occurrences in the function as they are the potentially hoistable
candidates. This approach allows us to hoist instructions to a basic block with
more than two successors. The $\chi$ node in a basic block $B$ has CHIArgs with
three entries: a) the instruction (I) CHIArg refers to, b) the basic block $Dest$
which is post-dominated by the basic block containing I, and c) the GVN of I.

@@

Limitations: This pass does not hoist fully redundant expressions because they
are already handled by GVN-PRE. It is advisable to run gvn-hoist before and
after gvn-pre because gvn-pre creates opportunities for more instructions to be
hoisted.

A $\chi$ keeps information about `congruent' values flowing out of a basic block.
It is similar to $\phi$ \cite{ssa} but in the inverse graph, and used for
trackig out-flowing values on each edge. Although it may be suitable for other
analyses/optimizations, for \GCM{}, it is computed only for instructions with
multiple occurrences in the CFG. An illustration of $\chi$ is shown in
Figure~\ref{fig:chi-intro}. The \GVN{} for both I1 and I2 is V, the $\chi$ node
will save the instruction as well as the edge where the value is flowing to.
Where $Arg1$, $Arg2$ are of type CHIArg. The $\chi$ for both I1 and I2 is inserted
in their PDF (A). For sinking we use the similar approach as \cite{ssapre} to
track values flowing into a basic block.

\begin{figure}
  \includegraphics[width=0.4\textwidth]{chi-example.pdf}
  \vspace*{-1.0cm}
\caption{I1, I2 are instructions in B, and C respectively, Arg1 = \{B, I1,
  V\} and Arg2 = \{C, I2, V\}, V is the value number for both I1, and I2}
\label{fig:chi-intro}
\end{figure}

When SSAPRE inserts $\Phi$ (expression PHIs), it does so at places where
(potentially) multiple expressions may merge. If a $\Phi$ has a missing entry,
that means the expression is partially available.  We apply a similar concept
for hoisting expression by working on inverted graph. We insert a $\chi$, to a
basic block with multiple successors. Then for each instruction, we register its
flow of value to its corresponding $\chi$ which can be found in its PDF. For
conciseness we only register values with multiple occurrences. Finally, we start
walking the CFG and remove the $\chi$ if it has a missing entry (see
Figure~\ref{fig:chi-intro2}), that means the expression is not anticipable in
the basic block having $\chi$. For the remaining $\chi$s anticipabilty of `values'
is ensured by construction.

\begin{figure}
  \includegraphics[width=0.4\textwidth]{chi-example2.pdf}
  \vspace*{-1.0cm}
\caption{I1, I2 are instructions in B, and C respectively, Arg1 = \{B, I1, V\}
  and Arg2 = \{C, I2, V\}, V is the value number for both I1, and I2. A has
  missing entry in $\chi$ so values are not anticipable}
\label{fig:chi-intro2}
\end{figure}

Before we consider factors affecting the legality of \GCM{}, we introduce the
concept of barriers.

@@

\subsubsection{Legality of movable instructions}
\label{subsec:legality}
Since the equality of movable candidates is purely based on the value numbers,
we also need to establish if hoisting them or sinking them would be legal. The
completeness of $\chi$ is a necessary condition for hoistability but not
sufficient. This is because congruence of value numbers only implies that they
compute same value if the inputs are same. So for instructions like memory
references, they may compute different values should the memory get modified
along the path. For that we need to check for intersecting side-effects along
the path.

For hoisting, we need to prove that on all execution paths from the hoisting
point to the end of the program the side effects appear in the same order. In
general, we cannot introduce a new computation along any path of execution
without checking for undefined behavior \cite{undef}.  For that, it is checked
that the side-effects of the computations (if any) do not intersect (alias) with
any side-effects between the instructions to be moved and their hoisting/sinking
point. It is also necessary to check if there are indirect branch targets, e.g.,
landing pad, case statements, goto labels, etc., along the path because it
becomes difficult to prove safety checks in those cases. In our current
implementation we discard candidates on those paths.

\subsubsection{Legality of hoisting scalars}
Scalars are the easiest to hoist because we do not have to analyze them for
aliasing memory references. As long as all the operands are available, or can be
made available by rematerialization, the scalar computations can be hoisted.

\subsubsection{Legality of hoisting memory references}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the address if possible.
Along the path, from current position of the load instruction backwards on the
control flow to the hoisting point, we check whether there are writes to memory
that may alias with the load, in which case the candidate is discarded. Memory
SSA allows efficient way of iterating over the use-def chains of memory
references on a factored graph.

For stores, we check the dependency requirements similar to the hoisting of
loads. We check that the operands of the store instruction are available at the
hoisting point, and there are no aliasing memory references along the hoisting
path.

\subsubsection{Legality of hoisting calls}
Call instructions can be divided into three categories: those calls equivalent
to purely scalar computations, calls reading from memory, and calls writing to
memory which is the most restrictive form.  Each category of call instructions
is handled as described for scalar, load, and store instructions.

Hoisting loads/stores across calls also requires precise analysis of all the
memory addresses accessed by the call. The current implementation being an
intraprocedural pass, cannot schedule aggressively across calls. In the presence
of pure calls, loads can be hoisted but stores can't. Also, if a call throws
exceptions, or if it may not return, nothing can be scheduled across that call.

\subsubsection{Legality of sinking expressions}
\label{subsec:legality-sink}
Sometimes, hoisting is not upward-safe \cite{click1995global}, e.g., if the
expressions are in a landing pad, sinking of those expressions
may reduce code size. Sinking may also reduce the register pressure in some
cases, e.g., when the use operands are not kills. For sinking, higher ranked
expressions would be sunk first. And it would be illegal to sink higher ranked
identical expressions if they are not fully anticipable in the common
post-dominator. For example:

B3: phi(c1, c2)
    phi(d1, d2)
\end{verbatim}

\subsection{Profitability check (Cost models)}
\label{subsec:cost-models}
TODO: Motivate why this is important.
After the legality checks have passed, we check if a \gcm{} is profitable.  That
takes into account the impact \gcm{} would have on various parameters that might
affect runtime performance, e.g., impact on live-range, gain in the code
size. The current implementation makes effort to not regress in performance
across popular benchmarks and at the same time to reduce code size as much as
possible. Following cost models are implemented:

@@

After hoisting, $a0$ and $a1$ are removed and a copy of $a0$ is placed in
$B0$. Since $b$ and $c$ are killed in both $a0$ and $a1$, hoisting the
expressions will reduce the register pressure because two registers will be
freed at the insertion point but only one register will be required to assign
the definition of hoisted instruction $a0$.

Ideally, it should be okay to hoist all the instructions and a later
live-range-splitting \cite{cooper1998live} pass should make the right decision
but the current live-range splitting pass of \LLVM{} is not making the optimal
decision and we have found performance regressions while hoisting aggressively.

There is a special case for hoisting load instructions when the hoist-point is
the predecessor basic-block for all the congruent loads. Even if the register
pressure would increase, we prefer to hoist loads in this case. The reason
being, that would make the loaded value available by the time it is used, and
livess would not increase by much because instruction is hoisted to immediate
predecessor.

@@

The compile time metrics of \GCM{} are listed in
Figure~\ref{tab:code-motion-metric}. The table lists the number of scalars,
loads, stores and calls hoisted as well as removed. For each category, the
number of instructions removed is greater or equal to the number of instructions
hoisted because each code-motion is performed only when at least one identical
computation is found. Loads are hoisted the most followed by scalars, stores and
calls in decreasing order.  This was the common trend in all our
experiments. One reason why loads are hoisted the most is the early execution of
this pass (before mem2reg pass which scalarizes some memory references) in the
\LLVM{} pass pipeline. Passes like mem2reg and instcombine might actually remove
those loads and the number of hoisted loads may change should the \GCM{} pass be
scheduled later. We can see a few sunk instructions even if \LLVM{} has a code
sinking and hoisting transforms in the SimplifyCFG pass which runs before
\GCM{}: this is because sinking and hoisting in SimplifyCFG have some rather
severe limitations to make the analysis and code transform very fast in order to
be able to run the pass several times.  Another reason is that instructions for
which dependencies are not directly available in the successor
(Section~\ref{subsec:legality-sink}) is not sunk for now because that would
require a look ahead on the ability to sink the dependencies. We plan to
implement this in near future.

@@

>>>>

%% Title information
\title[SSA GCM]{A fast algorithm for global code motion of congruent instructions}         %% [Short Title] is optional;

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a fast \gcm{} (\GCM{}) compiler optimization which schedules congruent
instructions across the program.  Not only \GCM{} saves code size, it exposes
redundancies as well, it exposes more instruction level parallelism in the
basic-block to which instructions are moved, and it enables other passes like
loop invariant motion to remove redundancies. The cost model to drive the code
motion is based on liveness analysis on \SSA{} representation such that the
(virtual) register pressure does not increase resulting in 2\% fewer spills on
the \SPEC{} benchmark suite.

@@

\section{A fast algorithm for code motion congruent computations}
Our algorithm factors out the reachability of values such that multiple queries
to find reachability of values are fast. This is based on finding the dominance
frontiers (DF) and the post-dominance frontiers (PDF) of the CFG. These points,
being structural properties of a graph, do not change during code-motion. A
post-dominance frontier is the basic block where anticipability (ANTIC) of an
instruction (I) is checked because it is the basic block on which the basic
block containing I is control dependent.

So we introduce a data structure ($\chi$ nodes) to keep track of values flowing
out of a basic block with more than one successor. This approach allows us to
hoist instructions to a basic block with more than two successors. The $\chi$
node in a basic block $B$ has CHIArgs with three entries: a) the instruction (I)
CHIArg refers to, b) the basic block $Dest$ which is post-dominated by the basic
block containing I, and c) the GVN of I.

@@

A $\chi$ keeps information about `congruent' values flowing out of a basic
block.  It is similar to $\phi$ \cite{ssa} but in the inverse graph, and used
for trackig out-flowing values on each edge. An illustration of $\chi$ is shown
in Figure~\ref{fig:chi-intro}. The \GVN{} for both I1 and I2 is V, the $\chi$
node will save the instruction as well as the edge where the value is flowing
to.  Where $Arg1$, $Arg2$ are of type CHIArg. The $\chi$ for both I1 and I2 is
inserted in their PDF (A).

\begin{figure}
  \includegraphics[scale=0.55]{chi-example.pdf}
  \vspace*{-1.0cm}
\caption{I1, I2 are instructions in B, and C respectively, Arg1 = \{B, I1,
  V\} and Arg2 = \{C, I2, V\}, V is the value number for both I1, and I2}
\label{fig:chi-intro}
\end{figure}

A similar idea has been used in SSAPRE \cite{ssapre} for partial redundancy
elimination in spare form. When SSAPRE inserts $\Phi$ (expression PHIs), it does
so at places where (potentially) multiple expressions may merge. If a $\Phi$ has
a missing entry, that means the expression is partially available. Subsequent
algorithms work on $\Phi$s to find out redundancies. We extend the idea for
\GCM{} by propagating the anticipability in both directions.

\begin{figure}
  \includegraphics[scale=0.55]{chi-example2.pdf}
  \vspace*{-1.0cm}
\caption{I1, I2 are instructions in B, and C respectively, Arg1 = \{B, I1, V\}
  and Arg2 = \{C, I2, V\}, V is the value number for both I1, and I2. A has
  missing entry in $\chi$ so V is not anticipable}
\label{fig:chi-intro2}
\end{figure}

For code hoisting we insert a $\chi$, to a basic block with multiple
successors. Then for each instruction, we register its flow of value to its
corresponding $\chi$ which can be found in its PDF. Finally, we start walking
the CFG and remove the $\chi$ if it has a missing entry (see
Figure~\ref{fig:chi-intro2}), that means the expression is not anticipable in
the basic block having incomplete $\chi$. For the remaining $\chi$s
anticipabilty of `values' is ensured by construction. For code-sinking we insert
$\Phi$s in basic blocks with multiple predecessors. Then for each instruction we
register its flow of value to its correponding $\Phi$ which can be found in its
DF.  Finally, we start walking the CFG and remove the $\Phi$ if it has a missing
entry (see Figure~\ref{fig:phi-intro}), that means the expression is not
available in the basic block having incomplete $\Phi$. For the remaining $\Phi$s
availability of `values' is ensured by construction.

\begin{figure}[h]
  \includegraphics[scale=0.55]{phi-example.pdf}
  \vspace*{-1.0cm}
  \caption{I1, I2 are instructions in B, and C respectively, Arg1 = \{B, I1, V\}
  and Arg2 = \{C, I2, V\}, V is the value number for both I1, and I2. A has
  missing entry in $\Phi$ so V is not available}
\label{fig:phi-intro}
\end{figure}

In order to limit the number of spurious $\chi$s and $\Phi$s,
\cite{ssapre-improv}, we insert them only for values with multiple occurrences
in the function as they are the interesting candidates for this \GCM{}. We
believe this factorization of anticipability/availability of expressions may be
suitable for faster implementation of other global (across basic blocks)
analyses/optimizations as well. Before we consider factors affecting the
legality of \GCM{}, we introduce the concept of barriers.

@@

\subsection{Legality of movable instructions}
\label{subsec:legality}
Since the equality of movable candidates is purely based on the value numbers,
we also need to establish if hoisting them or sinking them would be legal. The
completeness of $\chi$ and $\Phi$ are necessary for \GCM{} but not
sufficient. This is because congruence of value numbers only implies that they
compute same value if the inputs are same. So for instructions like memory
references, they may compute different values should the memory get modified
along the path. For that we need to check for intersecting side-effects along
the path.

For that, we need to prove that on all execution paths from the initial position
of an instruction to its destination, the side effects appear in the same
order. In general, we cannot introduce a new computation along any path of
execution without checking for undefined behavior \cite{undef}. It is also
necessary to check if there are indirect branch targets, e.g., landing pad, case
statements, goto labels, etc., along the path because it becomes difficult to
prove safety checks in those cases.

Scalars are the easiest to hoist because we only need to check for availability
of operands. For sinking the scalars we need to make sure that there are no uses
of the instruction along the path, For instructions with memory references like
loads, stores and calls, Memory SSA allows efficient way of iterating over the
use-def chains of memory references on a factored graph. Hoisting loads/stores
across calls also requires precise analysis of all the memory addresses accessed
by the call. The current implementation being an intraprocedural pass, cannot
schedule aggressively across calls. In the presence of pure calls, loads can be
hoisted but stores can't. Also, if a call throws exceptions, or if it may not
return, nothing can be scheduled across that call.

\subsection{Profitability check (Cost models)}
\label{subsec:cost-models}
After the legality checks have passed, we check if a \GCM{} is profitable.  That
takes into account the impact \GCM{} would have on various parameters that might
affect runtime performance, e.g., impact on live-range, gain in the code
size. The current implementation makes effort to not regress in performance but
at the same time to reduce code size as much as possible. Sometimes, hoisting is
not upward-safe \cite{click1995global}, e.g., if the expressions are in a
landing pad, sinking of those expressions may reduce code size. Following cost
models are implemented:

@@

After hoisting, $a0$ and $a1$ are removed and a copy of $a0$ is placed in
$B0$. Since $b$ and $c$ are killed in both $a0$ and $a1$, hoisting the
expressions will reduce the register pressure because two registers will be
freed at the insertion point but only one register will be required to assign
the definition of hoisted instruction $a0$. We have reduced performance
regressions with this cost model even if LLVM has a live-range splitting
\cite{cooper1998live} optimization.

Sinking may also reduce the register pressure in some cases, e.g., when the use
operands are not kills. For sinking, higher ranked expressions would be sunk
first. And it would be illegal to sink higher ranked identical expressions if
they are not fully anticipable in the common post-dominator. For example:

B3: PHIc1, c2)
    PHI(d1, d2)
\end{verbatim}

@@

The compile time metrics of \GCM{} are listed in
Figure~\ref{tab:code-motion-metric}. The table lists the number of scalars,
loads, stores and calls hoisted as well as removed. For each category, the
number of instructions removed is greater or equal to the number of instructions
hoisted because each code-motion is performed only when at least one identical
computation is found. Loads are hoisted the most followed by scalars, stores and
calls in decreasing order.  This was the common trend in all our
experiments. One reason why loads are hoisted the most is the early execution of
this pass (before mem2reg pass which scalarizes some memory references) in the
\LLVM{} pass pipeline. Passes like mem2reg and instcombine might actually remove
those loads and the number of hoisted loads may change should the \GCM{} pass be
scheduled later. We can see a few sunk instructions even if \LLVM{} has a code
sinking and hoisting transforms in the SimplifyCFG pass which runs before
\GCM{}: this is because sinking and hoisting in SimplifyCFG have some rather
severe limitations to make the analysis and code transform very fast in order to
be able to run the pass several times.  Another reason is that instructions for
which dependencies are not directly available in the successor
(Section~\ref{subsec:legality}) is not sunk for now because that would
require a look ahead on the ability to sink the dependencies. We plan to
implement this in near future.

@@

This pass does not hoist fully redundant expressions because they are already
handled by GVN-PRE. It is advisable to run gvn-hoist before and after gvn-pre
because gvn-pre creates opportunities for more instructions to be hoisted.
